---
title: "Predicting and Understanding what leads to High Knowledge Levels"
---

# Summary

In this study we look at how exam performance and time spent studying affect user knowledge on a specific application area. We see if we can build a regression model to accurately predict this user knowledge, and which if any of the two features are more important in building an accurate model.

# Introduction

A published paper on the development of intuitive knowledge classifiers collected data on undergraduate students in the Faculty of Technology in the Department of Computer Engineering at Gazi University in Turkey in October of 2009. It measured their knowledge on the subject of Electrical DC Machines using study time, study repetition and exam performance to determine their knowledge level. The data they released contains 6 total variables: the degree of study time for goal object materials, the degree of repetition number of user for goal object materials, the degree of study time of user for related object with goal object, the exam performance of user for related objects with goal object, the exam performance of user for goal objects, and finally the knowledge of the user. To simplify this data set for us to determine the knowledge level of the user, we decided to only use STG (the degree of study time for goal object materials), PEG (the exam performance of the user for goal objects), and UNS (the knowledge level of the user). In other words, from this data we can use study time, and exam performance to predict our own user’s knowledge level and test which is a better determining factor to knowledge level, study time or exam performance. We decided to use these two variables to determine user knowledge because we were curious on what mattered most for students’ knowledge level, study time or how they performed on the exam. This will also in turn let us understand if exams taken place actually evaluate what they are supposed to.<br/>

### Question: 
Which habit/result is more indicative of a students knowledge level: the time they spent studying or their actual exam results? <br/>

# Methods and Results

### Preliminary exploratory data analysis:

The dataset being used is User Knowledge Modeling Data Set and includes factors that measured the study time of more and less specific information for an exam as well as study repetition of this material. Exam performance on specific and less specific information was also measured. These factors were recorded on a scale of 0.00-1.00. These observations were then used to determine the target value of the knowledge level which was categorized with very low, low, middle and high. 

To begin answering our question, we began by loading the required packages.

```{r install library}
install.packages("rlang")
install.packages("tidymodels")
install.packages("kknn")
install.packages("knitr")
install.packages("rmarkdown")
install.packages("yaml")
install.packages("MASS")
install.packages("reshape2")
install.packages("ggplot2")
install.packages("caret")
```

```{r load library}
library(kknn)
library(tidyverse)
library(repr)
library(tidymodels)
library(readxl)
library(knitr)
library(rmarkdown)
library(MASS)
library(ggplot2)
library(caret)
```

Since the data set was an excel file and came from the web, we used the ```download.file``` function to download and convert the data from its original format to something we could work with and analyze. The excel also contained several sheets and was already split up into training and test data, which we used to create our training and test variables. The word format was inconsistent between sheets; the training data's user knowledge level was typed in lowercase and with underscores, and the test data's user knowledge level was typed in regular case and with spaces. So to ensure consistency between data, we converted the "Very Low" value into "very_low." The data was already standardized and so this step was not required here.

```{r load data}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00257/Data_User_Modeling_Dataset_Hamdi%20Tolga%20KAHRAMAN.xls"
data_file <- "data/data.xls"
download.file(url, data_file, mode = 'wb')
knowledge_train_data <- read_excel(data_file, 2)
knowledge_test_data <- read_excel(data_file, 3)
knowledge_test_data[knowledge_test_data == "Very Low"] <- "very_low"
```

With our data now loaded in, we select the specific columns (which include our predictors and classifier variables) from the data set. These columns include the User Knowledge Level (UNS, the classification variable), Study Time for Goal Object Materials (STG, a predictor), and Exam Performance for Goal Object (PEG, another predictor). Because the UNS column contains the classification groups, we mutated this from a string to an ordinal factor data type. We loaded the first couple of rows from each table to observe these changes. (See Table 1 and Table 2 below)

```{r create train/test split}

knowledge_train_data <- knowledge_train_data %>%
    mutate(UNS = factor(UNS, 
                        levels = c("very_low", "Low", "Middle", "High"), 
                        ordered = TRUE)) %>%
    dplyr::select(STG, PEG, UNS) %>%
    drop_na()

knowledge_test_data <- knowledge_test_data %>%
    mutate(UNS = factor(UNS, 
                        levels = c("very_low", "Low", "Middle", "High"), 
                        ordered = TRUE)) %>%
    dplyr::select(STG, PEG, UNS) %>%
    drop_na()

```

```{r Display Train}
head(knowledge_train_data)
```

*Table 1: First six rows of the Training Data <br/>
```{r Display Test}
head(knowledge_test_data)
```
*Table 2: First six rows of the Testing Data*

Next, we wanted to see the distribution of knowledge levels within the training and testing data, so we grouped the data by the knowledge level, found their counts, and then calculated the percent of each class. From Table 3, we can see that 9% of users have a very low knowledge level, 24% have a high knowledge level, 32% have a low knowledge level, and 34% have a middle knowledge level. Table 4 (testing data) on the other hand, shows that 18% of users have a very low knowledge level, 27% have a high knowledge level, 32% have a low knowledge level, and 23% have a middle knowledge level. We will have to keep these differences in mind during the randomization process and as we conduct further analysis.

```{r train data info}
num_obs <- nrow(knowledge_train_data)
know_percentage_train <- knowledge_train_data %>%
  group_by(UNS) %>%
  summarize(
    count = n(),
    percentage = n() / num_obs * 100 )
know_percentage_train
```
*Table 3: Count and Percent of User Knowledge Levels in the training data.*

```{r test data info}
num_obs2 <- nrow(knowledge_test_data)
know_percentage_test <- knowledge_test_data %>%
  group_by(UNS) %>%
  summarize(
    count = n(),
    percentage = n() / num_obs2 * 100 )
know_percentage_test
```
*Table 4: Count and Percent of User Knowledge Levels in the testing data.*

We still wanted to understand our data a bit more. To do so, we wrangled the training data a bit more to generate a table of the mean STG and PEG and the minimum and maximum values of the predictors for each knowledge level. This helped us get a sense of the boundaries for each class and the variance within each predictor (See Table 5 below).

```{r summarize data}
knowledge_train_summary <- knowledge_train_data %>%
    group_by(UNS) %>%
    summarize(count = n(), 
              mean_STG = mean(STG), 
              mean_PEG = mean(PEG),
             max_STG = max(STG), 
              max_PEG = max(PEG),
             min_STG = min(STG), 
              min_PEG = min(PEG))
knowledge_train_summary
```
*Table 5: Means, Minimums, and Maximums of Selected Variables in Training Data*

All these tables helped us understand the data but still required attentive interpretation. So we now created a visualization of the User Knowledge Level distribution based on the Exam Performance and Study Time variables. The plot below shows how the knowledge levels are stacked vertically like layers, where study time can vary from 0 to 1 for each class, but the exam performance imposes somewhat of a boundary on each category (See Figure 1 below).

```{r EDA create scatter plot}
options(repr.plot.width=10, repr.plot.height=6)
knowledge_train_plot <- knowledge_train_data %>%
    ggplot(aes(x=STG, y= PEG, colour = UNS)) +
        labs(x = "Degree of study time",
             y = "Exam performance",
            colour = 'Knowledge Level of Users') +
        geom_point() +
        theme(text = element_text(size = 20))
knowledge_train_plot
```
*Figure 1: Visualization of the distribution of Knowledge Levels based on Exam Performance and Study Time*

#### Main Analysis and Results

In this analysis we chose to use the machine learning method of ordinal logistic regression. Since this is a multi classification problem we could not use simple binary logistic regression. We begin by fitting our model and looking at the coefficients on each feature.

```{r Fit Model}
ordinal_model <- polr(UNS ~ STG + PEG, data = knowledge_train_data, Hess = TRUE)
```
```{r Summary of fit}
summary(ordinal_model)
```
*Table 6: Summary across included features in our fitted model*

From table 6 we see that exam performance has a coefficient nearly fifty times the degree of study indicating the feature is much more indicative of user knowledge. Now we have to see if this model performs well on the test data

```{r Predict on test data}
predicted_classes <- predict(ordinal_model, newdata = knowledge_test_data)
predicted_classes <- factor(predicted_classes, 
                            levels = levels(knowledge_test_data$UNS), 
                            ordered = TRUE)

# Model accuracy
mean(predicted_classes == knowledge_test_data$UNS)
```

Our model produces a mean accuracy of 82% which is promising. This means that really only using exam grade as a feature allows ordinal regression to achieve high accuracy across this complex problem.

Next we produce some visualizations to get a better sense of the data.

```{r Visualization of Confusion Matrix}
predicted_classes <- predict(ordinal_model, newdata = knowledge_test_data)

predicted_classes <- factor(predicted_classes, levels = levels(knowledge_test_data$UNS))

conf_matrix <- table(Predicted = predicted_classes, Actual = knowledge_test_data$UNS)

conf_matrix_df <- as.data.frame(conf_matrix)

ggplot(conf_matrix_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Confusion Matrix of Ordinal Logistic Regression",
       x = "Actual UNS Category",
       y = "Predicted UNS Category") +
  theme_minimal()
```
*Figure 2: Confusion Matrix*

This confusion matrix tells us where our model is predicting correctly and incorrectly, as well as the frequency to which it does that. We can see that across all possible classes it is fairly successful with some slight confusion being introduced when comparing the 'low' and 'very low' classes.

```{r Visualization of Feature Importance}
odds_ratios <- exp(coef(ordinal_model)) 

feature_importance <- data.frame(
  Feature = names(odds_ratios),
  Importance = odds_ratios
)

ggplot(feature_importance, aes(x = Feature, y = Importance, fill = Feature)) +
  geom_bar(stat = "identity") +
  labs(title = "Feature Importance in Ordinal Logistic Regression",
       y = "Odds Ratio (Feature Importance)") +
  theme_minimal() +
  scale_fill_viridis_d()
```
*Figure 3: Feature Importance*

In this figure we again look at how much importance our model gave to each of the features we included. As seen during fitting, exam grades is given the vast majority of the weighting and is practically the sole predictor for our model.

# Discussion

During this analysis we found a model that successfully predicts user knowledge separated by four distinct classes. We saw that when using degree of study time and exam performance as features, exam performance was by far the most useful predictor and dominated the model. This was an unexpected result in terms of the magnitude of importance. We thought that exam performance would probably be the most useful feature out of the two, but not by so much. This shows that the exams taken where this study was conducted is indeed a tell tale sign of user knowledge and that time spent studying, which may increase exam performance, does not tell us much about the users knowledge. 

The impacts this study could bring about are vast. By performing a similar result at other institutions it could be a way of confirming if their testing procedures are representative of real life skills. This is something every test maker hopes to achieve and could be substantiated by a similar result as in this study. 

In the future this could lead to questions regarding what else is indicative of user knowledge other than exam performance. Are there other indicators more important than exam performance?

We would also like to state that to improve this studies reliability we could make some or all of the following improvements. We could use cross validation to increase the confidence of our results, include more features to build a more robust and accurate model, and test out other regression techniques.